Notes on Validated Model Counting \\ Version of \today

### 2022-04-23 doc/mc-notes.tex
New example

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\pand}{\mathbin{\land_{\rm pv}}}
\newcommand{\por}{\mathbin{\lor_{\rm pa}}}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
%\newcommand{\tautology}{\top}
%\newcommand{\nil}{\bottom}
\newcommand{\tautology}{1}
\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

\newcommand{\interp}{\alpha}
\newcommand{\interpset}[1]{{\cal M}(#1)}
\newcommand{\ring}{{\cal Z}}
\newcommand{\cost}{\sigma}
\newcommand{\density}{\rho}
\newcommand{\hashset}{{\cal H}}
\newcommand{\fhash}{h}
\newcommand{\mcount}{\mu}

@@

Let $X = \{x_1, x_2, \ldots, x_n\}$ be a set of Boolean variables.  An
{\em assignment} is a function $\interp$ assigning Boolean values to
the variables: $\interp:X \rightarrow \{\nil, \tautology\}$.  We can
also view an assignment as a set of {\em literals} $\{l_1, l_2,
\ldots, l_n\}$, where each literal $l_i$ is either $x_i$ or
$\obar{x}_i$, corresponding to the assignments $\interp(x_i)$ = 1 or 0,
respectively.

@@

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment is then $1/2^{n}$, and the cost of a function is its
{\em density}, denoted $\density(f)$.  That is, the density of $f$ satisfies
$0 \leq \density(f) \leq 1$.  It
is the fraction of assignments for which the $f$
evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
$f$ can be scaled by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as the metric, rather than the number of models, has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.  As a
variant, some other forms of weighted model counting can be supported by
assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
\leq 1$ for every variable $x_i$.

{\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
and let $\hashset$ be the set of functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

\begin{lemma}[Complementation]
\label{lemma:complementation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f)$.
\end{lemma}

@@

@@

We introduce the CRAT clausal proof framework for generating a
checkable proof of the valuation of a Boolean formula, given in
conjunctive normal form, according to a specified cost function.  A
CRAT proof provides two forms of information: 1) a {\em schema} for computing a cost
function using the ring operations, and 2) a {\em proof} that this schema
accurately characterizes the cost function for the input formula.
A schema can be either {\em function independent}, meaning that it holds
for any cost function, or {\em function dependent}, meaning that
it holds for only for a specific cost function.

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item As with LRAT, a clause can be added as long as either 1) it is
  blocked, or 2) it satisfies the RUP/RAT property with respect to a supplied sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and to an abstraction rule.
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RUP/RAT proof
  that the two arguments are mutually exclusive.
\item The {\em abstraction} operation allows an extension
  variable to serve as an alias for one of more other 
  variables.  Use of this rule renders the schema function dependent.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item Unlike LRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[7mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP/RAT clause \\
     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RUP/RAT clause \\
    \midrule
        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
        & {\tt a} & $V \; L$         & $V^{+}$ {\tt 0}  & First application of abstraction \\ 
        & {\tt a} & $V \; L$         &                  & Subsequent applications of abstraction \\ 
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via
blocked-clause and RUP/RAT rules.  The hints portion of a
blocked-clause addition lists all earlier clauses containing
the negated version of the pivot literal, with the clause IDs negated.  The hints portion of a
resolution addition must contain a sequence of clause IDs such that
the added clause is RUP/RAT with respect to these clauses.  Clause
deletion requires an explicit justification that the deleted clauses
is RUP/RAT with respect to other clauses.

The second set of proof rules is unique to the CRAT format.  Each of
these indicates the addition of an extension variable.  For each case,
the rule must be followed by a sequence of blocked-clause additions.
A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
will represent the product $l_1 \pand l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \land l_2)$.
Literals $l_1$ and $l_2$ must have disjoint dependency sets.

A sum rule of the form ${\tt s}\;v\;l_1\;l_2$ indicates that $v$ will
represent the disjunction $l_1 \por l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \lor l_2)$.
The rule also contains a sequence of clause IDs such that the clause
$\obar{l}_1 \lor \obar{l}_2$ is RUP/RAT with respect to the sequence.

An abstraction rule of the form ${\tt a}\;v\;l$ indicates that
variable $v$ will be an alias for literal $l$.  The syntax differs
according to whether or not this is the first use of $v$ as an
alias.  The first instance must include a list of variables that is a
(possibly improper) superset of the dependency set for $l$.  It must
be followed by a sequence of blocked-clause additions encoding the
formula $v \leftrightarrow l$.  Subsequent instances do not
include an explicit dependency set, but $\depend(l)$ must be a subset of
those declared for $v$.  Furthermore, there are no blocked-clause
additions for these cases.

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion preserves
equivalence.  Starting with the set of input clauses, it produces a
sequence of steps that both add and delete clauses.  Each addition
must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause consisting of some variable or
its complement.  Except for trivial cases, the final variable will be
an extension variable.  The sequence of extension operations define
the schema for computing the cost function.  That is, a value for the
cost function is computed for every literal according to the rules of
Lemmas \ref{lemma:complementation}--\ref{lemma:disjunction}.  For
every abstraction step, all literals having the indicated extension
variable as their alias must evaluate to the same value.  The computed
cost is then the value computed for the literal in the final unit
clause.

\subsection{Example}

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Logical Form   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Evaluation Form\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$}
\label{fig:c3:schema}
\end{figure}

@@

As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
x_3$, represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations, as shown by the
logical form of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
correspond to $\pand$ operations and squares to $\por$ operations.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.

@@

>>>>

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\pand}{\mathbin{\land_{\sf v}}}
\newcommand{\por}{\mathbin{\lor_{\sf a}}}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
%\newcommand{\tautology}{\top}
%\newcommand{\nil}{\bottom}
\newcommand{\tautology}{1}
\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}
\newcommand{\pite}{\mbox{\it ITE}_{\sf v}}
\newcommand{\oneminus}{{\sim}}

@@

\newcommand{\interp}{\alpha}
\newcommand{\uinterp}{{\cal U}}
\newcommand{\interpset}[1]{{\cal M}(#1)}
\newcommand{\ring}{{\cal Z}}
\newcommand{\cost}{\sigma}
\newcommand{\density}{\rho}
\newcommand{\hashset}{{\cal H}}
\newcommand{\fhash}{h}
\newcommand{\mcount}{\mu}

@@

Let $X = \{x_1, x_2, \ldots, x_n\}$ be a set of Boolean variables.  An
{\em assignment} is a function $\interp$ assigning Boolean values to
the variables: $\interp:X \rightarrow \{\nil, \tautology\}$.  We can
also view an assignment as a set of {\em literals} $\{l_1, l_2,
\ldots, l_n\}$, where each literal $l_i$ is either $x_i$ or
$\obar{x}_i$, corresponding to the assignments $\interp(x_i)$ = 1 or 0,
respectively.  We denote the set of all assignments over these variables as $\uinterp$.

@@

For ring elemement $a$, we use the notation $\oneminus a$ to denote $1 - a$.

{\bf Example 1}: Let $\ring$ be the set of rational numbers of the form $a\cdot 2^b$ where $a$ and $b$ are integers.
Let $\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment is then $1/2^{n}$, and the cost of a function is its
{\em density}, denoted $\density(f)$.  That is, the density of $f$ satisfies
$0 \leq \density(f) \leq 1$.  It
is the fraction of assignments for which the $f$
evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
$f$ can be scaled by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as the metric, rather than the number of models, has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.

{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
{\em weight} $w_i$ for each variable $x_i$ such that $0 \leq w_i \leq
1$ and let $\cost(x_i) = w_i$.  This can support a restricted version of
{\em weighted} model counting.  The key restrictions are: 1) the
weight of an assignment equals the product of the weights of its
literals, and 2) the weight of a variable $x_i$ and its complement
$\obar{x}_i$ sum to 1.

{\bf Example 3}: Let $\ring$ be a field with $|\ring| \geq 2n$,
and let $\hashset$ be the set of functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

\begin{lemma}[Complementation]
\label{lemma:complementation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = \oneminus \cost(f)$.
\end{lemma}

@@

We will also find it useful to introduce a restricted version of the
if-then-else operator, denoted $\pite$.  That is, for functions $f_i$,
$f_t$, and $f_e$, we define $\pite(f_i, f_t, f_e) = (f_i \pand f_t)
\por (\neg f_i \pand f_e)$.  The two arguments to the disjunction are
guaranteed to satisfy the constraints for an assignment-partitioned
disjunction, since first argument can only be satisfied by assignments
for which $f_i$ evaluates to $\tautology$, while the second can only
be satisfied by assignments for which $f_i$ evaluates to $\nil$.  The
only restriction of the $\pite$ operator is that
$\depend(f_i) \cap \depend(f_t) = \emptyset$ and
$\depend(f_i) \cap \depend(f_e) = \emptyset$.

\section{Separable Schemas}

\begin{table}
  \caption{Recursive Definition of Separable Schemas}
  \label{tab:schema}
A) Primitive Operations\\
\begin{center}
\begin{tabular}{cccc}
\toprule
  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
\bottomrule
\end{tabular}
\end{center}
B) Derived Operations\\
\begin{center}
\begin{tabular}{cc}
\toprule
  $S$           & Definition \\
\midrule
\makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
\bottomrule
\end{tabular}
\end{center}
\end{table}

A {\em separable schema} is a directed acyclic graph representing a
Boolean formula where the only allowed operations are $\neg$, $\pand$,
$\por$ and $\pite$.  We use a graph representation to allow sharing of
subformulas.  The set of schemas over a set of variables $\{x_1, x_2,
\ldots, x_n\}$ can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.  We
divide the operations into {\em basic} ones that are fundamental to
the representation, and {\em derived} ones that can be constructed
from basic operations.  The latter category consists of just the $\pite$ operation.

A key property of a Boolean formula represented by separable schema
$S$ is that, for any separable cost function $\cost$, the cost of the
formula $\cost(S)$ can be computed with a linear number of ring
operations.

\begin{table}
  \caption{Normalization Rules}
  \label{tab:normalize}
  \begin{center}
  \begin{tabular}{ccccccc}
    \makebox[20mm]{$\neg \nil$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\tautology$} & \makebox[15mm]{} &
    \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
    $\neg \neg S$ & $\rightarrow$ & $S$ \\
    $S \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand S$ & $\rightarrow$ & $\nil$ \\
    $S \pand \tautology$ & $\rightarrow$ & $S$ && $\tautology \pand S$ & $\rightarrow$ & $S$ \\
    $S \por \nil$ & $\rightarrow$ & $S$ && $\nil \por S$ & $\rightarrow$ & $S$ \\
    $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
    $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
    $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
    $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
    $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
    $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
    $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
    $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
    $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
    $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a separable schema.  These are mostly
straightforward---eliminating extra negations and removing constant
terms.  The most interesting are the two bottom rules, were we make
use of DeMorgan's Laws to simplify an $\pite$ operation when one of
the arguments is $\tautology$.  We will make use of this
transformation when using BDDs to convert a set of clauses into a
separable schema.

@@

The CRAT clausal proof framework provides a means for creating a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to some separable schema.  This provides
a framework for adding proof generation to model counting programs.

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item As with LRAT, a clause can be added as long as either 1) it is
  blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ or $\pite$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RAT proof
  that the two arguments are mutually exclusive.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item A CRAT proof must show that the schema is logically equivalent
  to the input formula, not just that they are equisatisfiable.
  Therefore, each deletion step must also be show to be equivalence
  preserving, either because the clause is blocked or it follows from
  remaining clauses by the RAT property.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RAT clause \\
     & {\tt db} & $C$             & ${\tt -}C^{+}$  {\tt 0} & Delete blocked clause \\
     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RAT clause \\
    \midrule
        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via blocked-clause and
RAT rules.  As described below, however, blocked clauses can only be added
to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
of a blocked-clause addition lists all earlier clauses containing the
negated version of the pivot literal, with the clause IDs negated.
The hints portion of a RAT addition must contain a sequence of clause
IDs such that the added clause is RAT with respect to these clauses.
Clause deletion requires an explicit justification that the deleted
clauses is RAT with respect to other clauses.

The second set of proof rules is unique to the CRAT format.  Each of
these indicates the addition of an extension variable.  For each case,
the rule must be followed by a sequence of blocked-clause additions
providing the defining clauses for the extension variable.

A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
will represent the product $l_1 \pand l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \land l_2)$.
Literals $l_1$ and $l_2$ must have disjoint dependency sets.

A sum rule of the form ${\tt s}\;v\;l_1\;l_2$ indicates that $v$ will
represent the disjunction $l_1 \por l_2$.  The blocked clause
additions must encode the formula $v \leftrightarrow (l_1 \lor l_2)$.
The rule also contains a sequence of clause IDs such that the clause
$\obar{l}_1 \lor \obar{l}_2$ is RAT with respect to the sequence.

An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
clause additions must encode the formula
$v \leftrightarrow \ite((l_1,l_2, l_3)$.
The dependency set for $l_1$ must be disjoint from those
of $l_2$ and $l_3$.

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion preserves
equivalence.  Starting with the set of input clauses, it produces a
sequence of steps that both add and delete clauses.  Each addition
must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause consisting of some variable or
its complement.  Except for trivial cases, the final literal will be
an extension variable or its complement.  That variable will indicate the root node of the schema.
Working from that root backward, the schema can be extracted from the CRAT file.

\subsection{Example 1}

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$ and its Cost Computation}
\label{fig:c3:schema}
\end{figure}

@@

As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
x_3$, represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations.  Figure~\ref{fig:c3:schema}A shows one such decomposition.
Circles in this schema
correspond to $\pand$ operations and squares to $\por$ operations.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.

@@

\subsection{Example 2}

\begin{figure}
  \centering{
  \begin{tabular}{llllll}
    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
\midrule
    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause $C_1$}\\

    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
            & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
    {\tt 6} & {\tt ab}  & {\tt 5 1 -4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
    {\tt 8} & {\tt ab}  & {\tt -5 4 0} & {\tt -6 0}    & & \\
    {\tt 9} & {\tt ar}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
            & {\tt dr}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
            & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
    {\tt 10} & {\tt ab}  & {\tt 6 -1 2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
    {\tt 11} & {\tt ab}  & {\tt -6 1 0}    & {\tt -10 0} & & \\
    {\tt 12} & {\tt ab}  & {\tt -6 -2 0}    & {\tt -10 0} & & \\ 
    {\tt 13} & {\tt ar} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
            & {\tt dr}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
    {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -5 0} &  \\  
    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -5 0} & &  \\
    {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
             & {\tt dr}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
             & {\tt dr}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
  \end{tabular}
}  
  \caption{CRAT Proof Steps for Formula $(x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2$}
  \label{fig:p2:crat}
\end{figure}

As a more complex example, consider the Boolean formula given by the
conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more direct encoding of the problem.
The complement of the formula can be written in DNF as
$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
the two conjuncts can be formed using $\pand$ operations, and their
sum can be formed using a $\por$ operation.  In this example, we first
form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
$C_2$.  These are asserted as unit clauses in proof lines 9 and 13,
allowing the input clauses to be deleted.  Then we generate $t_3$ as
the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
Based on this, we can delete the unit clauses for terms $t_1$ and
$t_2$.  Term $t_3$ then becomes the root of the schematic representation.
